Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-28 appendix.tex
revised
major revision in process because of barker & latter

In the global disk with $\Gamma=1$ we have
\begin{align}\label{dkappa2}
  \frac{\p\kappa^2}{\p z} = 4 \frac{\p\Omega^2}{\p z} + r\frac{\p}{\p
    r}\frac{\p\Omega^2}{\p z} = -
   \frac{\p\ln\rho}{\p z}\frac{qc_s^2}{r^2} \left(
  \frac{3z^2}{r^2+z^2}-1\right) \equiv - \frac{\p\ln\rho}{\p z}\frac{qc_s^2}{r^2}F(r,z).
\end{align}
% For the local problem we may then write
% \begin{align}
%   \frac{dD}{dz}  = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2}F(z),
% \end{align}
The function $F$ varies monotonically from $F=-1$ at $z=0$ to $F\to2$
as $|z|\to\infty $, so $|F|=O(1)$. 
% For the local problem we set
% $dD/dz = \p\kappa^2/\p z$. 
% This function varies from $F=q$ at $z=0$ to $F\to
% 3+q$ as $|z|\to\infty$. Since we consider $|q|=O(1)$, this impies
% $|F|=O(1)$ as well.

@@

\section{Linear problem in the isothermal limit}\label{iso_discuss}  
Here we examine in more detail the special case of isothermal
perturbations ($\beta\equiv 0$) for vertically isothermal disks 
($\Gamma=1$) in the VGSB framework.   
%Isothermal perturbations correspond to $\beta\equiv0$ ($\chi=1$) and 
In this case Eq. \ref{ode_Q} becomes   
\begin{align} 
  Q = W 
\end{align}
(i.e. $\delta P = c_s^2\delta \rho$), and so $  \ii\hat{\sigma}c_s\delta v_z = W^\prime $
from Eq. \ref{lin_vz}. For isothermal perturbations it is in fact
simpler to work with an equation for $W$ by substituting $Q=W$ into
Eq. \ref{ode_w}. In this case, we shall not yet make the low frequency
approximation, but first make the nearly Keplerian approximation. We
obtain, in terms of dimensionless variables, 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \frac{\ii \epsilon q \hat{k} f(\zhat)
      }{1-\hat{\sigma}^2} \right]W^\prime +
  \hat{\sigma}^2\left(1+\frac{\hat{k}^2}{1-\hat{\sigma}^2}\right)W,\label{iso_ode0} 
\end{align}
where for discussion purposes we have defined $f(\zhat)$ such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = \epsilon^2q f(\hat{z})\Omega_k^2.
\end{align}
By comparing Eq. \ref{fz_shear} with Eq. \ref{vertical_shear_ex}, we
see that $f(\zhat)= 
\zhat\left(1+\epsilon^2\zhat^2\right)^{-3/2}$. More generally, though,
$f$ may be regarded as a representation of the vertical
shear profile. Physically, we expect there is a maximum value of 
$|d\Omega/dz|$, the existence of which should limit the growth rate of
the VSI,  as remarked in \S\ref{vshear_def}. We explicitly demonstrate
this below.

@@

\subsection{Explicit solutions in the thin-disk limit}\label{iso_explicit}
Here we assume a thin disk ($\epsilon\ll1$) so that $\ln\rho\simeq
-\zhat^2/2$ and $f(\zhat)\simeq \zhat$. However, we do not assume 
low frequency from the onset. Then Eq. \ref{iso_ode0} becomes 
\begin{align}\label{iso_ode3}
  W^{\prime\prime} - \left(1 + \frac{\ii q\epsilon \hat{k}}{1-\hat{\sigma}^2}\right)\hat{z}W^\prime  +
  \hat{\sigma}^2\left(1+\frac{\hat{k}^2}{1-\hat{\sigma}^2}\right)W = 0.
\end{align}
We remark that taking the low-frequency limit and considering
$\hat{k}^2\gg 1$, Eq. \ref{iso_ode3} becomes equivalent to Eq. 39 in
\cite{nelson13}, although we have taken a different route.

@@

\subsubsection{Stability without vertical shear}\label{stable_novshear}
In the absence of vertical shear $q=0$, Eq. \ref{simple_growth} can be
written as \begin{align}
  \hat{\sigma}^2\khat^2 =
  \left(L-\hat{\sigma}^2\right)\left(1-\hat{\sigma}^2\right), \label{iso_disp_full}
\end{align}
which is just the dispersion relation for axisymmetric isothermal waves in a
vertically isothermal disk
\citep[e.g.][]{okazaki87,takeuchi98,tanaka02,zhang06,ogilvie13,barker14}. In 
this case the solutions are Hermite polynomials, $W\propto
\He_L(\zhat)$.  The eigenfrequency $\hat{\sigma} = \hat{\omega}$ is real and the disk
is stable. The low frequency branch of Eq. \ref{iso_disp_full} are 
inertial waves \citep{balbus03}. For 
$|\hat{\omega}|\ll 1$ and $L\geq 1$ the dispersion relation is
$\hat{\omega}^2\khat^2 = L$, or $\hat{\omega}\propto \khat^{-1}$ for
fixed $L$.  This inverse relation has been qualitatively
observed in numerical simulations of \cite{stoll14}.

\subsubsection{Instability with vertical shear}
The VSI corresponds to unstable inertial waves. This is readily
obtained for large $\khat^2$ by balancing
the last two terms of Eq. \ref{simple_growth}  to obtain the low
frequency branch. Then 
\begin{align}
  \hat{\sigma}^2 \simeq L\left(\frac{1+\ii q \epsilon
       \hat{k}}{L+1+\hat{k}^2}\right).\label{simple_growth2}
\end{align}
This signifies instability for $L\geq1$ since we can choose the sign
of the square root to make $\hat{\nu} = \imag\hat{\sigma}>0$.  These
are the low frequency unstable modes seen in
Fig. \ref{compare_modes_iso_kx10} and
Fig. \ref{compare_modes_iso_kx30} , for which $\nu\propto\omega$.

>>>>

In the global disk with $\Gamma=1$ we have
\begin{align}\label{dkappa2}
  \frac{\p\kappa^2}{\p z} = 4 \frac{\p\Omega^2}{\p z} + r\frac{\p}{\p
    r}\frac{\p\Omega^2}{\p z} = -
   \frac{\p\ln\rho}{\p z}\frac{qc_s^2}{r^2} \left(
  \frac{3z^2}{r^2+z^2}-1\right) \equiv - \frac{\p\ln\rho}{\p z}\frac{qc_s^2}{r^2}F(r,z).
\end{align}
% For the local problem we may then write
% \begin{align}
%   \frac{dD}{dz}  = - \frac{d\ln\rho}{dz}\frac{qc_s^2}{r^2}F(z),
% \end{align}
The function $F$ increases monotonically from $F=-1$ at $z=0$ to $F\to2$
as $|z|\to\infty $, so $|F|=O(1)$. 
% For the local problem we set
% $dD/dz = \p\kappa^2/\p z$. 
% This function varies from $F=q$ at $z=0$ to $F\to
% 3+q$ as $|z|\to\infty$. Since we consider $|q|=O(1)$, this impies
% $|F|=O(1)$ as well.

@@

\section{Linear problem in the isothermal limit}\label{iso_discuss}  
Here we summarize some results for isothermal
perturbations ($\beta\equiv 0$) in vertically isothermal disks 
($\Gamma=1$) in the VGSB framework. In this case Eq. \ref{ode_Q}
becomes    
\begin{align} 
  Q = W 
\end{align}
(i.e. $\delta P = c_s^2\delta \rho$), and so $  \ii\hat{\sigma}c_s\delta v_z = W^\prime $
from Eq. \ref{lin_vz}. For isothermal perturbations it is in fact
simpler to work with an equation for $W$ by substituting $Q=W$ into
Eq. \ref{ode_w}. In this case, we shall not yet make the low frequency
approximation, but first make the nearly Keplerian approximation. We
obtain, in terms of dimensionless variables, 
\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \frac{\ii \epsilon q \hat{k} f(\zhat)
      }{1-\hat{\sigma}^2} \right]W^\prime +
  \hat{\sigma}^2\left(1+\frac{\hat{k}^2}{1-\hat{\sigma}^2}\right)W,\label{iso_ode0} 
\end{align}
where for discussion purposes we have defined $f(\zhat)$ such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = \epsilon^2q f(\hat{z})\Omega_k^2.
\end{align}
By comparing Eq. \ref{fz_shear} with Eq. \ref{vertical_shear_ex}, we
see that $f(\zhat)= 
\zhat\left(1+\epsilon^2\zhat^2\right)^{-3/2}$. More generally, though,
$f$ may be regarded as a representation of the vertical
shear profile. Physically, we expect there is a maximum value of 
$|d\Omega/dz|$, the existence of which should limit the growth rate of
the VSI,  as remarked in \S\ref{vshear_def}. We explicitly demonstrate
this below.

@@

\subsection{Explicit solutions in the thin-disk limit}\label{iso_explicit}
Here we assume a thin disk ($\epsilon\ll1$) so that $\ln\rho\simeq
-\zhat^2/2$ and $f(\zhat)\simeq \zhat$. However, we do not assume 
low frequency from the onset. Then Eq. \ref{iso_ode0} becomes 
\begin{align}\label{iso_ode3}
  W^{\prime\prime} - \left(1 + \frac{\ii q\epsilon \hat{k}}{1-\hat{\sigma}^2}\right)\hat{z}W^\prime  +
  \hat{\sigma}^2\left(1+\frac{\hat{k}^2}{1-\hat{\sigma}^2}\right)W = 0.
\end{align}
We remark that taking the low-frequency limit and considering
$\hat{k}^2\gg 1$, Eq. \ref{iso_ode3} becomes equivalent to Eq. 39 in
\cite{nelson13} or Eq. 28 in \cite{barker15}, although we have taken a
different route.

@@

Note that we have applied a regularity condition at infinity, since
the vertically isothermal disk has no surface. If vertical boundaries
are imposed at finite height, as done in numerical calculations, then
the above solution needs to be modified to match the desired boundary
conditions. This enables the `surface modes' seen in numerical
calculations \citep{barker15}.

\subsubsection{Stability without vertical shear}\label{stable_novshear}
In the absence of vertical shear $q=0$, Eq. \ref{simple_growth} can be
written as \begin{align}
  \hat{\sigma}^2\khat^2 =
  \left(L-\hat{\sigma}^2\right)\left(1-\hat{\sigma}^2\right), \label{iso_disp_full}
\end{align}
which is just the dispersion relation for axisymmetric isothermal waves in a
vertically isothermal disk
\citep[e.g.][]{okazaki87,takeuchi98,tanaka02,zhang06,ogilvie13,barker14,barker15}. In 
this case the solutions are Hermite polynomials, $W\propto
\He_L(\zhat)$.  The eigenfrequency $\hat{\sigma} = \hat{\omega}$ is real and the disk
is stable. The low frequency branch of Eq. \ref{iso_disp_full} are 
inertial waves \citep{balbus03}. For 
$|\hat{\omega}|\ll 1$ and $L\geq 1$ the dispersion relation is
$\hat{\omega}^2\khat^2 = L$, or $\hat{\omega}\propto \khat^{-1}$ for
fixed $L$.  This inverse relation has been qualitatively
observed in numerical simulations of \cite{stoll14}.

\subsubsection{Instability with vertical shear}
The VSI corresponds to unstable inertial waves \citep{barker15}. This
is readily obtained for large $\khat^2$ by balancing
the last two terms of Eq. \ref{simple_growth} to give the low
frequency branch. Then 
\begin{align}
  \hat{\sigma}^2 \simeq L\left(\frac{1+\ii q \epsilon
       \hat{k}}{L+1+\hat{k}^2}\right), \label{simple_growth2}
\end{align}
which is equivalent to Eq. 34 of \cite{barker15} in the limit
$\khat^2\gg L$. This signifies instability for
$L\geq1$ since we can choose the sign 
of the square root to make $\hat{\nu} = \imag\hat{\sigma}>0$.  These
are the low frequency unstable modes seen in
Fig. \ref{compare_modes_iso_kx10} and
Fig. \ref{compare_modes_iso_kx30}, for which $\nu\propto\omega$.

### 2015-03-28 linear_adia.tex
major revision in process because of barker & latter

\subsection{Isothermal perturbations}\label{iso_pert}
Isothermal perturbations correspond to $\beta = 0$ or $\gamma=1$ so 
that $\chi = 1$. This case has been investigated numerically
by \cite{nelson13} and \cite{mcnally14}.

In Appendix \ref{iso_discuss}, we give an analytical discussion of the
linear problem for isothermal perturbations. There, we establish more general
results that do not require the thin disk approximation, namely 
a necessary condition for instability (there must be vertical shear)
and an upper bound to the instability growth rate (it is limited by
the maximum vertical shear rate in the disk). We also develop 
solutions in the thin disk limit. The 
methods used in Appendix \ref{iso_discuss} and results obtained
therein motivates the solution procedure employed in the following
sections.

>>>>

\subsection{Isothermal perturbations}\label{iso_pert}
Isothermal perturbations correspond to $\beta = 0$ or $\gamma=1$ so 
that $\chi = 1$. This case has been investigated by \cite{nelson13},
\cite{mcnally14} and \cite{barker15}.

In Appendix \ref{iso_discuss}, we collect some useful results for
isothermal  perturbations. This includes a necessary condition for
instability (there must be vertical shear) and an upper bound to the
instability growth rate (it is limited by the maximum vertical shear
rate in the disk). We also develop solutions in the thin disk
limit. The methods used in Appendix \ref{iso_discuss} and results obtained
therein motivates the solution procedure employed in the following
sections, and are provided for ease of reference. We refer the reader
to \cite{barker15} for a more detailed treatment of isothermal 
perturbations.

### 2015-03-28 linear_iso.tex
revised
major revision in process because of barker & latter

\section{Analytical discussion of vertically isothermal
  disks in the framework of the vertically global shearing
  box}\label{analytical}  
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the radially-local
approximation. 
In order to facilitate a fully analytic discussion, we first introduce a
number of approximations below that allows a simple governing equation
to be derived and compared with previous works.  However, all of the 
approximations described below are relaxed in our
numerical study.

@@

In deriving Eq. \ref{vertiso_gov}, the vertical dependence of
$D=\kappa^2(z)-\sigma^2$ was ignored because the replacement
$D\to\Omega_k^2$ was made before eliminating variables in favor of
$\delta v_z$. In Appendix \ref{adia_improve} we show that making this
replacement after eliminating variables, which involves $dD/dz$,
introduces terms $O(\epsilon^2)$. Since we are interested in 
thin disks ($\epsilon\ll 1 $), e.g. protoplaneary disks where 
$\epsilon\lesssim 0.1$ \citep{chiang10}, these terms introduce
unnecessary complexity for the present discussion.

>>>>

\section{Analytical discussion of vertically isothermal
  disks in the framework of the vertically global shearing
  box}\label{analytical}  
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the radially-local
approximation. 
In order to facilitate a fully analytic discussion, we first introduce a
number of approximations below that allows a simple governing equation
to be derived and compared with previous works. Similar approximations
are employed by \cite{nelson13} and \cite{barker15}. However, all of
the approximations described below are relaxed in our numerical study.

%vertically isothermal disk lacks a surface

@@

In deriving Eq. \ref{vertiso_gov}, the vertical dependence of
$D=\kappa^2(z)-\sigma^2$ was ignored because the replacement
$D\to\Omega_k^2$ was made before eliminating variables in favor of
$\delta v_z$. In Appendix \ref{adia_improve} we show that making this
replacement after eliminating variables, which involves $dD/dz$,
introduces terms $O(\epsilon^2)$. Since we are interested in 
thin disks ($\epsilon\ll 1 $), e.g. protoplanetary disks where 
$\epsilon\lesssim 0.1$ \citep{chiang10}, these terms introduce
unnecessary complexity for the present discussion.

### 2015-03-28 paper.tex
major revision in process because of barker & latter

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in stably stratified disks.  
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of accretion disks in which the disk  
  temperature $T$ is relaxed its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response of the disk by
  varying $\beta$. We find increasing $\beta$ even slightly away from 
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. 
  We apply our linear calculations  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. We find the MMSN is unstable to the VSI from $\sim 5$AU to $\sim 50$AU with  
  a characteristic growth time of $\sim 30$ local orbital 
  periods, and the dominant VSI mode in the MMSN is the `fundamental
  corrugation mode' which causes the disk column to oscillate
  vertically. We suggest the VSI is dynamically important in the outer
  parts of protoplanetary disks. 
\end{abstract}

@@

The VSI can be regarded as the disk analog of the
Goldreich-Schubert-Fricke instability 
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, enables
perturbations with large vertical length-scales (relative to radial 
length-scales) to become unstable. \cite{umurhan13} gives a simple
interpretation for the VSI: an exchange of fluid parcels while
conserving their angular momenta individually, may result in a new
configuration of lower energy, if the exchange occurs over a
sufficiently large vertical displacement. In practice, the VSI
corresponds to unstable inertial waves.  
%inertial waves 
%history

The VSI has been studied analytically in local approximation
\citep{urpin98,urpin03}.  Numerical simulations of the  
VSI have been performed by \cite{arlt04}, and more recently by 
\cite{nelson13} and \cite{stoll14}. In these new simulations the authors
indeed observe the growth of vertically global but radially localized
perturbations when the equilibrium rotation varied with height.

An important finding from these works is that the VSI requires a short
thermal timescale: perturbations need to be close to
isothermal, unless the disk is neutrally stratified in the vertical
direction. This is because vertical motions, required for the VSI, are stabilized by
buoyancy if the disk is sub-adiabatically stratified for
convective stability. When the thermal timescale $t_c$ is parametrized
in units of the orbital period $P_\mathrm{orb}$, \cite{nelson13} find that
$t_c\lesssim 0.01P_\mathrm{orb}$ was needed to observe instability. That is, the
disk temperature essentially needs to remain unchanged during its
evolution. This was further supported with radiation hydrodynamic
simulations by \cite{stoll14}, who find the VSI is only sustained with
the inclusion of heating from stellar irradiation, which resulted in
an effective thermal timescale $t_c\ll P_\mathrm{orb}$.

From these studies it is clear that the VSI is sensitive to
the thermodynamic response of the disk if it is stably
stratified. Since the latter is usually assumed for typical disk models,  
it is important to understand and quantify said thermodynamic dependence,
for example on disk parameters, in order to evaluate the relevance of
the VSI in realistic disks. Should a given disk become turbulent
through the VSI, it must first pass through the
linear instability. Thus, understanding the thermodynamic dependence
of the linear VSI is essential.

In this work we study the linear VSI in radially local,
vertically global models of astrophysical disks. We include an energy
equation with a source term that allows the thermodynamic response of
perturbations to be varied smoothly between isothermal and adiabatic. Our
formulation generalizes previous linear calculations of the 
VSI carried out by \cite{nelson13} and \cite{mcnally14} which assume
isothermal perturbations and therefore did not include an energy
equation. Our models are also an extension of earlier 
analysis by \cite{urpin98} and \cite{urpin03}, which accounts for thermal
evolution, but considered vertically localized disturbances. Our 
vertically global approach is necessary to study disturbances with
vertical wavelengths comparable to or larger than the disk scale
height. Such disturbances were  found by \cite{nelson13} and
\cite{stoll14} to dominate their numerical simulations.  We thus focus
on vertically global perturbations which, in fact, we find to be the
dominant mode in protoplanetary disks with a realistic estimate for
the thermodynamic timescales.

>>>>

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in stably stratified disks.  
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of accretion disks in which the disk  
  temperature $T$ is relaxed its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response of the disk by
  varying $\beta$. We find that in a stably stratified disk,
  increasing $\beta$ even slightly away from    
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. 
  We apply our linear calculations  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. We find the MMSN is unstable to the VSI from $\sim 5$AU to $\sim 50$AU with  
  a characteristic growth time of $\sim 30$ local orbital 
  periods, and the dominant VSI mode in the MMSN is the `fundamental
  corrugation mode' which causes the disk column to oscillate
  vertically. We suggest the VSI is dynamically important in the outer
  parts of protoplanetary disks. 
\end{abstract}

@@

The VSI can be regarded as the disk analog of the
Goldreich-Schubert-Fricke instability 
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, can render
inertial waves unstable. \cite{barker15} gives a simple 
interpretation of the VSI: in an axisymmetric inviscid disk, the 
exchange of fluid parcels vertically, while conserving their specific angular
momentum, can result in a new configuration of lower energy if the
distribution of specific angular momentum varies with $z$. The VSI is
characterized by perturbations with large vertical length-scales
(relative to radial length-scales).

Early analysis of the VSI in radially and vertically local 
models of accretion disks \citep{urpin98,urpin03} showed that the VSI
requires a short thermal timescale so that perturbations evolve 
isothermally, unless the disk is neutrally stratified in the vertical
direction. This is because vertical motions associated with the VSI
are stabilized by buoyancy forces if the disk is sub-adiabatically
stratified (convectively stable). A sufficiently short thermal
timescale renders buoyancy ineffective.

The need for a short thermal timescale in stably stratified disks has
been demonstrated in numerical simulations of the VSI
\citep{nelson13,stoll14}. When the  thermal timescale $t_c$ is
parametrized in units of the orbital period  $P_\mathrm{orb}$,
\cite{nelson13} find that $t_c\lesssim  0.01P_\mathrm{orb}$ was needed
to observe the VSI. That is, the  disk temperature needs to remain
effectively unchanged. Radiation hydrodynamic simulations performed by
\cite{stoll14} also showed that the VSI is only sustained with the
inclusion of heating from stellar irradiation, which results in an
effective thermal timescale $t_c\ll P_\mathrm{orb}$ that maintains a
fixed temperature profile.

Given that the VSI is generally sensitive to disk thermodynamics,  
it is important to understand and quantify this dependence, for
example on disk parameters, in order to evaluate the 
relevance of the VSI to realistic disks, in which one may expect
non-zero vertical entropy gradients and thermal timescales. Should a
given disk become turbulent through the VSI, it must first pass
through the linear instability. Thus, understanding the thermodynamic
dependence of the  linear VSI is essential.

In this work we study the linear VSI in radially local, vertically
global models of astrophysical disks. This approach has
been adopted in recent studies of the linear VSI
\citep{nelson13,mcnally14,barker15}. However, these studies consider 
isothermal perturbations, for which the thermal timescale is formally
zero, and/or neutrally stratified disks in which the stabilizing
effect of bouyancy is absent. We account for both effects to study the
VSI in more general disk models. To do so, we examine 
sub-adiabatically stratified disks and include an energy equation 
with a source term that allows the thermodynamic response of 
perturbations to be varied smoothly between isothermal and 
adiabatic.

% Our models are also an extension of earlier 
% analysis by \cite{urpin98} and \cite{urpin03}, which accounts for thermal
% evolution, but considered vertically localized disturbances.
% Our 
% vertically global approach is necessary to study disturbances with
% vertical wavelengths comparable to or larger than the disk scale
% height. Such disturbances were found by \cite{nelson13} and
% \cite{stoll14} to dominate their numerical simulations. We thus focus
% on vertically global perturbations which, in fact, we find to be the
% dominant mode in protoplanetary disks with a realistic estimate for
% the thermodynamic timescales.
